\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid

### 2015-02-19 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Finished my pass of the paper

There have been periodic efforts to reduce or remove the need for parameter tuning.
\cite{rechenberg:1973:es} introduced self-adaptive parameters, in which parameter values
were included in each solution's genome and themselves underwent evolution. This allowed search
to optimize some of its own parameters resulting in a reduced the need for expert tuning.
\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.

@@

Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
for any problem, the order in which problem variables appear in the genome can be changed
without changing the behavior of the search. The fitness can also be manipulated in any fashion
as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
will not impact search efficiency.

@@

Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
such that none of the single bit neighbors are fitness improvements, search stops. Thus
this algorithm requires a restart mechanism to solve multimodal problems. We have
chosen here to naively restart search from a random solution whenever a local optima is found. This
ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.

@@

By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
the population size, and uses a fixed user specified size for each generation.

@@

Other algorithms have proposed using multiple concurrent populations.
\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
upward. This allows for continuous integration of diversity as the lowest population is reseeded
with random solutions. However, this method resulted in increased parameterization as not only was
a population size required, but also parameters for how frequently generations advanced between levels and
how many total levels to have were added. \cite{harik:1999:parameterlessga} used multiple independent
populations of different sizes as a method for removing the population size parameter, but
doing was provably less efficient than using an optimal population size as no information is shared
between the populations.

@@

Perhaps the most common model for generating random rugged landscapes is the NK model. An
NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
other genes in the genome. This fitness is specified using a randomly generated table of
fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
if epistasis is set such that each gene depends on the $K$ directly following it
in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.

@@

\section{Comparison Algorithm Parameter Tuning}
\label{sec-tuning}

@@

While four of the six algorithms in our experiments do not require any user specified parameters,
hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
to determine the optimal population size for each problem size.
Extended by \cite{goldman:2012:ltga}, this method iteratively
doubles the population size until some success criteria is met and then performs bisection
between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
As bisection can make infinitely large population sizes, any run which had not found the global
optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.

@@

\subsection{Quantitative Comparison}
Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
Pairwise comparison of LTGA and P3 on the largest problem size using the Mann-Whitney~U test results in
$p < 10^{-5}$ for all problems.

\subsection{Local Search}
The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
sizes. This is especially true for the three randomly generated problem problem classes. These problems
may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
better than any other algorithm. We believe this is
because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
flipping entire traps.
This ability also leads $1+(\lambda, \lambda)$
to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.

@@

\subsection{Quantitative Comparison}
Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
were the next best, outperforming P3 on 27, 20, and 18 sample points. Random Restart Hill Climbing almost
never outperformed P3, doing so only 9 times.

@@

Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by
reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
Neighbor NK.
Even when using a reduced population size for LTGA, P3 still achieved a fitness at
least as high as LTGA at 80\% of intervals.

@@

This behavior is rooted in the peculiar nature
of this landscape.
After local search, all traps in all solutions have a total
number of 1 bits equal to \{0, 1, 3, 5, 7\} as these correspond to the local
optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
two bit plateaus, and as a result solutions in the second level generally do not
contain the lowest fitness local optima (5 bits set) and the third
level has very few traps set to the next worst local optima (3 bits set). As a result,
solutions which reach the third level can only be improved by replacing the deceptive
local optima (0 and 1 bits set) with the global optimum (7 bits set).
The global optimum is very rare in the population, and with 8 ways to represent local optima
linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
meaning few solutions will be added to the fourth level. Solutions that do improve by
definition must have a higher frequency of optimal trap settings, meaning level four's model
will be more accurate and donations are more likely to contain optimal trap values.
Thus the level size and crossover success rates increase after contracting around level four.

@@

Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
constant factor, while for the three problem classes P3's improvement increased with problem size. This
quality extends to intermediate points during evolution, with P3 generally reaching at least as high
of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
is some evidence that even when compared on wall clock time, P3 performs on par with
the best comparison techniques. All of these achievements are made without any problem specific parameter
tuning, making P3 easier to apply to new domains than its two closets competitors in quality.

>>>>

There have been periodic efforts to reduce or remove the need for parameter tuning.
\cite{rechenberg:1973:es} introduced self-adaptive parameters, in which parameter values
were included in each solution's genome and themselves underwent evolution. This allowed search
to optimize some of its own parameters, resulting in a reduced need for expert tuning.
\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.

@@

Only hBOA and LTGA require any parameters, with each of these only requiring a population
size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
for any problem, the order in which problem variables appear in the genome can be changed
without changing the behavior of the search. The fitness can also be manipulated in any fashion
as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
will not impact search efficiency.

@@

Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
such that none of the single bit neighbors are fitness improvements, search stops. Thus
this algorithm requires a restart mechanism to solve multimodal problems. We have
chosen here to na\"{i}vely restart search from a random solution whenever a local optima is found. This
ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.

@@

By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
the population size, relying instead on a fixed user specified size.

@@

Other algorithms have proposed using multiple concurrent populations.
\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
upward. This allows for continuous integration of diversity as the lowest population is reseeded
with random solutions. However, this method resulted in increased parameterization as not only was
a population size required, but also new parameters for how frequently generations advanced between levels and
how many total levels to have. \cite{harik:1999:parameterlessga} used multiple independent
populations of different sizes as a method for removing the population size parameter, but
doing was provably less efficient than using an optimal population size as no information is shared
between the populations.

@@

Perhaps the most common model for generating random rugged landscapes is the NK model. An
NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
other genes in the genome. This fitness is specified using a randomly generated table of
fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
if epistasis is set such that each gene depends on the $K$ directly following it
in the genome, the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.

@@

\section{Comparison Algorithm Parameter Tuning}
\label{sec-tuning}
While four of the six algorithms in our experiments do not require any user specified parameters,
hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
to determine the optimal population size for each problem size.
Extended by \cite{goldman:2012:ltga}, this method iteratively
doubles the population size until some success criteria is met and then performs bisection
between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
As bisection can make infinitely large population sizes, any run which had not found the global
optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.

@@

@@

\subsection{Quantitative Comparison}
Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
according to the binomial test.
Pairwise comparison of LTGA and P3 on the largest problem size using the Mann-Whitney~U test results in
$p < 10^{-5}$ for all problems.

\subsection{Local Search}
The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
sizes. This is especially true for the three randomly generated problem classes. These problems
may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
better than any other algorithm. We believe this is
because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
flipping entire traps.
This ability also leads $1+(\lambda, \lambda)$
to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.

@@

\subsection{Quantitative Comparison}
Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
were the next best, outperforming P3 on 27, 20, and 18 sample points, respectively. Random Restart Hill Climbing almost
never outperformed P3, doing so only 9 times.
The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-9}$
according to the binomial test.

@@

Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by
reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
Neighbor NK.
Even when using a reduced population size for LTGA, P3 still achieved a fitness at
least as high as LTGA at 80\% of intervals.
The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
according to the binomial test.

@@

This behavior is rooted in the peculiar nature
of this landscape.
After local search, all traps in all solutions have a total
number of 1 bits equal to either 0, 1, 3, 5, or 7 as these correspond to the local
optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
two bit plateaus, and as a result solutions in the second level generally do not
contain the lowest fitness local optima (5 bits set) and the third
level has very few traps set to the next worst local optima (3 bits set). As a result,
solutions which reach the third level can only be improved by replacing the deceptive
local optima (0 and 1 bits set) with the global optimum (7 bits set).
The global optimum is very rare in the population, and with 8 ways to represent local optima
linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
meaning few solutions will be added to the fourth level. Solutions that do improve by
definition must have a higher frequency of optimal trap settings, meaning level four's model
will be more accurate and donations are more likely to contain optimal trap values.
Thus the level size and crossover success rates increase after contracting around level four.

@@

Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
constant factor, while for the three randomly generated problem classes P3's improvement increased with problem size. This
quality extends to intermediate points during evolution, with P3 generally reaching at least as high
of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
is some evidence that even when compared on wall clock time, P3 performs on par with
the best comparison techniques. All of these achievements are made without any problem specific parameter
tuning, making P3 easier to apply to new domains than its two closets competitors in quality.

### 2015-02-19 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/reviews.tex
Finished my pass of the paper

We believe that the comments can be broken into three classes:
\begin{itemize}
\item Errors in the paper that need correcting
\item Minor fixes/modifications that we can implement.
\item Items of disagreement
\end{itemize}

@@

This was indeed an error on our part. Unfortunately the source code was not made available to us, despite a request,
and the code was not yet released to the public until after the paper was submitted. The fix,
a two line modification, does not affect any of the evaluation counts but does indeed modify our
timed comparisons (wall clock time experiments). Subsequently, after moving the entropy calculation,
we were able to reach timing parity with Dr. Bosman's implementation. We have rerun all of our timing results
and updated our findings accordingly. In general we find that for single instance problems,
optimally tuned LTGA %(where we of course do not count the evaluations or timing necessary to find the optimal population size)
finds the global optimum using less seconds than P3. However, on randomly generated problem classes,
P3's reduction in evaluations overcomes its increased CPU usage per evaluation.
We have updated our figures and discussion to reflect these changes.

@@

We agree, though our overall point is more general in that we are trying to get the community to rethink
\textbf{all} parameter settings. In the original submission Section 9, Section 10.2,
Figure 11, and Figure 13a explicitly discuss the benefits we believe P3 achieves due
to side stepping the fixed size generational scheme.
We also had previously mentioned it in the third paragraph of the conclusion.
Nonetheless, we have now added some additional references and discussion to the introduction to indicate its importance.

@@

We agree that it is important for neutral drift to occur during mixing. Our original submission included that feature,
as stated in Figure 3 line 6 and near the end of Section 2.5. We believe the source of this confusion is that P3
requires strict fitness improvement when adding solutions to the pyramid, as stated in Figure 4 line 9 and near the beginning
of Section 3.

\item
\begin{quote}
The abstract states that P3 requires "less wall clock seconds" and "is more effective than existing state-of-the-art techniques".
\end{quote}

@@

\item
\begin{quote}
"Doerr et al. (2013) presented the first genetic algorithm to provably show the advantage of performing crossover on simple landscapes."
I understand some people have a very restricted view on provably showing something but this statement is arguably wrong.
\end{quote}

We did not intend to be offensive and have cleaned up that text.

@@

Replaced "updating" with "rebuilding"

\item
\begin{quote}
"Furthermore, unlike LTGA all solutions in the population are used to generate the model, not just the winners of a binary tournament".
This is not a major design choice, this only influences the selection pressure. Actually, the most recent papers report all experiments with LTGA using the entire population instead of a set of selected solutions.
\end{quote}

@@

Though we have stated previously that adding to the pyramid triggers modelbuilding, we have added text to the figure to restate it.

@@

\item
\begin{quote}
 "This function modifies the results of Equation 4 to include plateaus of size s, introducing an exponential number of local optima in each trap."
Exponential number ? Clarify with a small example of Deceptive Step Trap.
\end{quote}

We have added the following text for clarity. 
With $k=7$ and $s=2$, as used in our experiments, all traps with 0, 1, 3, 5, and 7 bits set are local optima.
This means that half of all ways to set the trap are 1 bit local optima.
More generally, the number of local optima grows at $\Theta(2^{k-1})$.

@@

\item
\begin{quote}
"To ensure each problem is challenging we chose a clause to variable ratio of 4.27"
Reference to support this number.
\end{quote}

@@

\item
\begin{quote}
 "An important note is that each donation may range in size from a single bit up to N-1, with smaller donations far more likely to result in repeated attempts. As such this may cause some super-linear growth in P3, but its unlikely to be very high."
Speculative ?
\end{quote}

@@

\item
\begin{quote}
" In total these runs took over 6 computing years to complete."
\end{quote}

@@

This is covered in the original text, Section 2.4 and 2.5. However, to review,
the run time bottleneck for LTGA is extracting pairwise frequency information which
requires $O(\mu N^2)$ time per generation. This happens approximately once every $\mu N$ evaluations,
meaning the time per evaluation is $(\mu N^2) / (\mu N) = O(N)$.
In P3, both entropy extraction and model rebuilding are $O(N^2)$ and must be done every $N$ evaluations,
meaning the time per evaluation is $N^2 / N = O(N)$. Therefore, excluding caveats mentioned in the paper,
the two techniques should be within a constant factor runtime of each other.
This relationship in general holds true for our updated experiments.

\item
\begin{quote}
 "There is some evidence that even when compared on wall clock time, P3 performs at least as well as the best comparison techniques"
\end{quote}

Fixed based on the algorithmic fix. Now ``on par'' instead of ``better.''

\item
\begin{quote}
"While asymptotically linear"   Linear in what ?
\end{quote}

@@

We have also added statistical tests to data from Figure 6 to Section 6.1.

\item
\begin{quote}
Last sentence of section 1, penultimate clause doesn't have  verb in it. I suggest you rephrase as " ...; analyse the algorithm complexity; 
\end{quote}

@@

\item
\begin{quote}
Section 8.2.1 uses the phrase "significantly worse".  This suggests that you have some specific confidence in the results according to some hypothesis testing method.  Please clarify or rephrase.
\end{quote}

@@

However, adding initial local search as a tunable parameter to hboa and LTGA when partial evaluations are not available is not part
of existing work for either method. Doing so would also provide those techniques
with even more problem specific configuration. We welcome future work by other researchers exploring the possible effects of combining
local search with their techniques.

>>>>

We believe that the comments can be broken into three classes:
\begin{itemize}
\item Errors in the paper that have been corrected.
\item Minor fixes/modifications that we have addressed.
\item Items of disagreement
\end{itemize}

@@

This was indeed an error on our part. Unfortunately the source code was not made available to us, despite a request,
and the code was not released to the public until after the paper was submitted. The fix,
a two line modification, does not affect any of the evaluation counts but does indeed modify our
wall clock time experiments. Subsequently, after moving the entropy calculation,
we were able to reach timing parity with Dr. Bosman's implementation. We have rerun all of our timing results
and updated our findings accordingly. In general we find that for single instance problems,
optimally tuned LTGA %(where we of course do not count the evaluations or timing necessary to find the optimal population size)
finds the global optimum using less seconds than P3. However, on randomly generated problem classes,
P3's reduction in evaluations overcomes its increased CPU usage per evaluation.
We have updated our figures and discussion to reflect these changes.

@@

We agree, though our overall point is more general in that we are trying to get the community to rethink
\textbf{all} parameter settings. In the original submission Section 9, Section 10.2,
Figure 11, and Figure 13a explicitly discuss the benefits we believe P3 achieves due
to side stepping the fixed size generational scheme.
We also had previously mentioned it in the third paragraph of the conclusion.
Nonetheless, we have now added some additional references and discussion to the introduction to elevate its importance.

@@

We agree that it is important for neutral drift to occur during mixing. Our original submission included that feature,
as stated in Figure 3 line 6 and near the end of Section 2.5. We believe the source of the reviewer's confusion is that P3
requires strict fitness improvement when adding solutions to the pyramid, as stated in Figure 4 line 9 and near the beginning
of Section 3.

\item
\begin{quote}
The abstract states that P3 requires ``less wall clock seconds'' and ``is more effective than existing state-of-the-art techniques''.
\end{quote}

@@

\item
\begin{quote}
``Doerr et al. (2013) presented the first genetic algorithm to provably show the advantage of performing crossover on simple landscapes.''
I understand some people have a very restricted view on provably showing something but this statement is arguably wrong.
\end{quote}

We have made the claim more precise to better reflect their actual accomplishment.

\item
\begin{quote}
typo:  ``in a reduced the need for''
\end{quote}

@@

Replaced ``updating'' with ``rebuilding''

\item
\begin{quote}
``Furthermore, unlike LTGA all solutions in the population are used to generate the model, not just the winners of a binary tournament''.
This is not a major design choice, this only influences the selection pressure. Actually, the most recent papers report all experiments with LTGA using the entire population instead of a set of selected solutions.
\end{quote}

@@

Though we have stated previously that adding to the pyramid triggers model building, we have added text to the figure to restate it.

@@

\item
\begin{quote}
 ``This function modifies the results of Equation 4 to include plateaus of size s, introducing an exponential number of local optima in each trap.''
Exponential number ? Clarify with a small example of Deceptive Step Trap.
\end{quote}

We have added the following text for clarity. 
``With $k=7$ and $s=2$, as used in our experiments, all traps with 0, 1, 3, 5, and 7 bits set are local optima.
This means that half of all ways to set the trap are 1 bit local optima.
More generally, the number of local optima grows at $\Theta(2^{k-1})$.''

@@

\item
\begin{quote}
``To ensure each problem is challenging we chose a clause to variable ratio of 4.27''
Reference to support this number.
\end{quote}

@@

\item
\begin{quote}
``An important note is that each donation may range in size from a single bit up to N-1, with smaller donations far more likely to result in repeated attempts. As such this may cause some super-linear growth in P3, but its unlikely to be very high.''
Speculative ?
\end{quote}

@@

\item
\begin{quote}
``In total these runs took over 6 computing years to complete.''
\end{quote}

@@

This is covered in the original text, Section 2.4 and 2.5. However, to review,
the runtime bottleneck for LTGA is extracting pairwise frequency information which
requires $O(\mu N^2)$ time per generation. This happens approximately once every $\mu N$ evaluations,
meaning the time per evaluation is $(\mu N^2) / (\mu N) = O(N)$.
In P3, both entropy extraction and model rebuilding are $O(N^2)$ and must be done every $N$ evaluations,
meaning the time per evaluation is $N^2 / N = O(N)$. Therefore, excluding caveats mentioned in the paper,
the two techniques should be within a constant factor runtime of each other.
This relationship in general holds true for our updated experiments.

\item
\begin{quote}
``There is some evidence that even when compared on wall clock time, P3 performs at least as well as the best comparison techniques''
\end{quote}

Fixed based on the corrected code. Now ``on par'' instead of ``better.''

\item
\begin{quote}
``While asymptotically linear''   Linear in what ?
\end{quote}

@@

We have also added statistical tests of data from Figure 6, Figure 7, and Figure 12 to Section 6.1, Section 7.1, and Section 9.2.

\item
\begin{quote}
Last sentence of section 1, penultimate clause doesn't have  verb in it. I suggest you rephrase as `` ...; analyse the algorithm complexity;''
\end{quote}

@@

\item
\begin{quote}
Section 8.2.1 uses the phrase ``significantly worse''.  This suggests that you have some specific confidence in the results according to some hypothesis testing method.  Please clarify or rephrase.
\end{quote}

@@

However, adding initial local search as a tunable parameter to hboa and LTGA when partial evaluations are not available is not part
of existing work for either method. Doing so would also provide those techniques
with even more problem specific configuration. We welcome future work by other researchers exploring the possible effects of combining
local search with their techniques in this fashion.
